Reproducibility of research during COVID-19: examining the case of population density and the basic reproductive rate from the perspective of spatial analysis

### 2021-09-16 R0-Density-Reanalysis/Response-to-Reviewers-v2.Rmd

---
title: "Response to Reviewers"
output:
  pdf_document: default
date: "`r format(Sys.time(), '%d %B, %Y')`"
bibliography: bibliography.bib
nocite: |
  @Gordon1989gasoline, @Newman1989gasoline
---

Dear Editor:

I would like to thank you and the two anonymous reviewers for the attention paid to my submission entitled "Reproducibility of research during COVID-19: examining the case of population density and the basic reproductive rate from the perspective of spatial analysis" (GEAN-06-21-053).

The reviewers are generally positive about the contents of the paper, and there are no comments requiring additional analytical work. The most serious comment, from Reviewer 2, concerns the _tone_ and _presentation_ of the paper. You ask that I revise the paper to ensure that I am fair to a junior researcher (Reese Sy). This is a reasonable request. In this letter I respond to the thoughtful feedback I received and how I revised the paper in response.

I would like to assure the reviewers and you that my intention in using their work as an example was to constructively engage with Sy, White, and Nichols (SWN). In fact, even before submission to GA I had already taken steps to ensure that I was not acting like a complete jackass and a bully. I hope that my responses will persuade you of this.

Sincerely,

The author

<!--
Note to self: check this on how to create a document with the differences: https://tex.stackexchange.com/questions/165543/using-latexdiff-in-miktex
-->

\newpage

## Reviewer 1

The study aims to demonstrate the value of reproducible research, which involves making code and data publicly available. A strength of the manuscript is the expansion of relevant concerns with the SWN model, from the perspective of spatial analysis, and extensive and detailed justification for every model specification choice in the reanalysis. The inclusion of a RMarkdown file also guides the reader in understanding the models used in the reanalysis.

I have no major comments that can improve the study, as I believe the study is analytically sound and well-written. The only minor comment I have is the length of the background of the association of density and R0, which could be condensed further. Overall, the study is a good example of a manuscript that demonstrates the value of reproducible research, and would be an outstanding contribution to the scientific literature.

> \color{blue}{Many thanks for your comments and positive assessment of the paper. I lightly edited the background section for conciseness, while trying to keep the paper as self-contained as possible.}

\newpage

## Reviewer 2

The stated goal of this article is to promote reproducible research. This is done by reviewing another paper (SWN) that followed the standards advocated by this paper.

> \color{blue}{Thank you for this comment. In fact, I reviewed at least 22 papers for this research. Of these, SWN was almost alone in being reproducible. Furthermore, of those that were reproducible, SWN was the only one that I thought I could engage with in an interesting way. The papers that I reviewed, along with notes about their reproducibility, are in the table below.}

Paper | Reproducibility Notes
---   | ---
@Ahmad2020association | Code not shared, data are provided but data pre-processing is not documented
@Amadu2021assessing | Code not shared, data provenance is documented, data are not shared
@Bhadra2021impact | Code and data not available; only data provenance is documented
@Brandtner2021creatures | Code is available on demand, data are shared
@Cruz2020exploring | Code is shared, sources of data are documented, data are not actually provided
@Feng2020spread | Code not shared, data provenance is documented, data are not shared
@Feyman2020effectiveness | Code and data are provided
@Fielding2020social | Code not shared, data provenance is documented, data are not actually provided
@Hamidi2020longitudinal |  Code and data not available
@Hamidi2020density |  Code and data not available
@Inbaraj2021seroprevalence | Code not shared, data are not shared
@Lee2020human | Code not provided, data sources are documented but data are not shared (privacy & confidentiality)
@Noury2021how | Code not shared, data are shared
@Pequeno2020air | Raw data supplied in supplemental files, code not available
@Roy2020factors | Code not shared, data are provided and data processing is documented
@Skorka2020macroecology | Code and data are shared but data processing is not documented, 
@Souris2020covid | Code not provided, data sources are documented but data are not shared
@Stephens2021impact | Code and data are shared
@Sy2021population | Code and data are shared
@Wang2021transmission | Code not shared, data are available
@White2020state | Code and data are in repository
@Wong2020spreading | Code not provided, data are shared but data pre-processing is not documented

My summary of the paper: 1) It is partly a new "COVID-19 paper" because it provides a new way to understand the population density hypothesis. 2) It is partly a "response paper" because it provides in-depth criticism of a recent paper. 3) It is partly an "advocacy paper" because it is clearly promoting open science. I will address these in order.

> \color{blue}{Thank you for this insightful summary, which I think captures quite well the different missions of the paper.}

1) The new models shown look sound and are an improvement over other models in this field. The density hypothesis is often discussed, so it is important to continuously improve our understanding of it.

> \color{blue}{Thank you for your positive assessment of the technical aspects of the paper.}

2) Since this will not be published in PLOS, the original authors will not have the opportunity to respond. It does not feel fair. The authors of the GA paper might argue, “this is not a response paper.” SWN seems to be a strange choice for a larger example. The first author is a graduate student. No one should be immune from scientific criticism (the selection bias criticism is valid), but choosing a generally good graduate-led paper (if I correctly understand the GA authors) feels wrong. If the GA authors chose the papers led by well-established authors such as Paez or Wong, I would not feel so queasy.

> \color{blue}{Thank you for the thoughtful response to this aspect of my paper. I would like to begin my response by noting that PLoS ONE desk-rejected my paper, which deprived me of an opportunity to engage in scientific discourse with SWN (apparently PLoS ONE does not think that science should be discursive). Even worse, by forcing me to take my paper elsewhere, PLoS ONE deprived SWN of the opportunity to respond to my paper in their journal of choice. This is, in my view, a very weird aspect of PLoS ONE's editorial policy. Secondly, I did not choose SWN for the example to pick on them, but rather because they were, as I noted above, almost the only example of reproducible research that I found in my research. I chose SWN partly of necessity, and partly as a homage, not because I particularly wished to criticize them.}

> \color{blue}{More generally, I fully agree with you and the editor that criticism should be professional, in particular when directed at the work of junior researchers. Graduate students should be nurtured, not crushed. I tried to be respectful of Sy, White, and Nichols, which is why I carefully checked myself to avoid any and all arguments that could be construed as ad hominem. I tried to engage with the science, not with them personally. I also made an effort to recognize that I, just like everyone else, have strengths and limitations as researcher; I don't pretend to have the last work on density and the spread of COVID, and recognize that others may be better equipped to analyze these data more competently than me.}

> \color{blue}{Incidentally, your comment brought to mind memories of Gordon and Richardson's (1989) response to Newman and Kenworthy's (1989) paper on Gasoline Consumption and Cities. In their response, Gordon and Richardson concluded by suggesting that "[p]erhaps Newman and Kenworthy would be well advised to seek out another planet, preferably unpopulated, where they can build their compact cities from scratch with solar-powered transit." I read this many years ago as a graduate student, and to the date I still find it shockingly unprofessional. I must admit that I lack the viciousness and ferocity of Gordon and Richardson. Rather, my aim is to be a good scientist, a decent professional, but if at all possible an even better human being. For this reason I took steps, from the moment I completed the first draft of the paper, to ensure that I was operating within the parameters of my moral compass:}

> \color{blue}{1. I submitted the paper to PLoS ONE so that, if accepted, it would be a companion to the original SWN paper. As I noted above, PLoS ONE issued a desk rejection within two days: apparently their editorial policy precludes "commentary"; this leads me to believe that a response from SWN would also not have been acceptable in PLoS ONE. Only after being rejected there, did I try a different journal for the paper.}

> \color{blue}{2. Immediately after I pushed the first version of the paper to my repository, thereby making it public, I communicated with Reese Sy, Laura White, and Brooke Nichols. This was my e-mail to them on May 21, 2021:}.

> \color{blue}{\textit{Dear Ms. Sy, Dr. White, and Dr. Nichols, I would like to express my appreciation for your recent work on population density and the spread of COVID-19. I came across your paper while doing a search for a review assignment, and it caught my attention given some of my earlier work on COVID-19. Your paper is a relatively rare example of reproducible research, and it allowed me to relatively quickly replicate your results. I submitted a paper to PLoS ONE that builds on your analyses, and I thought that it would be just courteous to share it with you. Please see: *repository*. The README page needs some tidying up (for some reason a figure is not displayed). The source document is an R Markdown document which, along with the pdf you can find in this folder: R0-Density-Reanalysis. I also packaged the data for ease of distribution, and the workflow for data preparation is documented in this folder: raw-data-processing). If you have any questions or comments about this, I would be happy to hear from you.}}

>\color{blue}{As you can see from the salutation, I was well-aware that Ms. Sy is a graduate student. This was the response from Prof. White to my email:}

> \color{blue}{\textit{Dear Dr. X, Thanks for sharing this. I enjoyed reading your paper and find your results really interesting. I am happy to see you addressing the selection bias issues and considering nonlinearities in the association between density and R0. As you noted, these were issues that we chose not to address, but I am happy to see this picked up on. I agree with your premise the making our research reproducible only leads to better science. Hopefully we all get a bit better at this going forward. Best regards, Laura}}

> \color{blue}{This was followed by a response from Prof. Nichols:}

> \color{blue}{\textit{Dear X, I would like to echo Laura’s words. I really enjoyed reading your work- and this is the type of thing we should do more often in science! Cheers, Brooke}}

> \color{blue}{Last but not least, this was the response from Ms. Sy:}

> \color{blue}{\textit{Dear X, I agree with Laura and Brooke - thank you for sharing this with us. I completely agree with what you wrote about reproducible research as vital to examine the same data from a variety of different perspectives. This strengthens scientific discourse! Have a wonderful day. Best, Reese}}

> \color{blue}{Why did I ask SWN to comment on a first version of the paper? First, I strongly felt it was the courteous and decent thing to do. And secondly, I think that science is better and more fun when it is done in a spirit of cooperation instead of antagonism. Their responses do not give a sense that they felt the paper was unfair to any of them as a team, let alone to Ms. Sy as an individual. Quite the opposite. I was very happy (not to mention relieved!) when their responses were as supportive as they were.}

> \color{blue}{3. In addition, when submitting the paper to PLoS ONE (and after that to Science of the Total Environment, Landscape and Urban Planning, and Geographical Analysis), I always recommended that at least one of the authors of SWN be invited to review the paper.}

> \color{blue}{It seems clear to me that PLoS ONE is not interested in responses to research, only in research. But does PLoS ONE's editorial policy really mean that SWN cannot respond? If fairness to SWN is important, the Editor of GA could easily invite them to respond to my paper in GA! Since PLoS ONE does not seem to care about this conversation, why shouldn't we take it to continue it elsewhere?}

3) A well-documented paper containing data and code is easier to replicate and improve than a paper without these items. This seems obvious. The paper possibly indicates to journal editors that response papers are more likely to be submitted under open science. It possibly forms a cautionary tale that authors need to be extra diligent if they make their work easily replicable. However, using a case study method, as the GA paper does, seems to hurt the original paper/authors more than promote open science.

> \color{blue}{I agree with you that this seems obvious. The reality is that, as I noted above, the examples of replicable research are still few and far between, and in the whole literature of density and COVID, SWN was in the minority. I think my message is only partially directed to editors: for example, PLoS ONE already has a data and code policy - and yet, it is so seldom followed, let alone enforced, as to be next to useless. My message, instead, is mainly directed to researchers: we need to do much better than the very lowly expectations of journals. I already explained that I did not choose SWN to pick on them, but rather because they were one of the precious few papers that were reproducible. My communications with them do not give the impression that they were hurt by this paper, but rather felt it was an encouraging product of open science. And, if in doubt, GA could always invite them to respond.}

I want to think the GA paper comes from a place that respects the SWN paper, but the implementation of the GA paper does not give that impression. However, I do like the improved COVID-19 density model.

>\color{blue}{I am sorry that you were left with the impression that I was not respectful of SWN. I have no reason to think that they themselves felt disrespected. In my revisions I have been extra-careful to ensure that I engage with the material in a respectful and professional way.}

Other suggestions

Schweinsberg et al. (2021) supports the view that different researchers reach different conclusions (https://doi.org/10.1016/j.obhdp.2021.02.003).

>\color{blue}{Thank you for suggesting this paper. I have included it in the discussion and references.}

Pages 4,5. I think the authors have swapped the meanings of "large" and "small" scales.

>\color{blue}{Thank you for pointing this out. The text has been revised to avoid confusion.}

The authors did not mention the time period studied.

>\color{blue}{SWN obtained publicly available daily COVID-19 cases among United States counties from the New York Times and for each county assumed that the exponential growth period began one week prior to the second daily increase in cases. We assumed that the period of exponential growth lasted approximately 18 days, as previous research have shown the COVID-19 exponential period to be around 20–24 days in New York City.}

I assume that the data for the GA paper comes from the Johns Hopkins database. This data is sometimes revised, so this particular data set is an example of citations being insufficient to achieve exact replicability.

>\color{blue}{The data were obtained from the New York Times. Since the New York Times uses GitHub for their data repository, data versioning is available. See: https://github.com/nytimes/covid-19-data.}

Please define the basic reproductive number for non-epidemiologist readers earlier in the paper.

>\color{blue}{Done.}

The first part of the discussion section contains additional results. The discussion seems to start on page 18, line 41.

>\color{blue}{I reorganized the discussion following this comment.}

These opinions on the SWN paper and its authors seem unnecessary.
Page 7, line 30. "reasonable modeling choice"
Page 7, line 40. "sensible controls"
Page 8, lines 34,35. "whether a graduate student or a seasoned scientist"
Page 18, lines 46-47. "required different technical skills than those available to SWN"

>\color{blue}{I appreciate your attention to these statements, which have been revised to make them more a matter of factly and less opinionated. Some statements were deleted altogether since they were unnecessary.}

Page 8, line 19. It seems that the authors call themselves "expert"s. This seems unnecessary.

>\color{blue}{This was more of a general statement rather than an exlusive reference to me, and has been changed to "independent researcher" to avoid implying expertise.}

Page 9, lines 49-52. There are many reasons to delete this footnote.

>\color{blue}{It is a fact that the paper was desk rejected by the journals for the reasons stated. The fact that one of them was PLoS ONE helps to explain why publishing this paper elsewhere is not unfair to SWN. I did delete the editorial comment to the effect that "[t]his does not inspire much confidence in the commitment of journals to reproducibility in research."}

Citation typos: page 6, line 44; page 6, line 51; page 7, line 9; page 7, line 20; page 8, line 51

>\color{blue}{Errors corrected.}

Typos
Page 9, line 20. "RWN" should be "SWN"
Page 11, line 21. "fourth" should be "fifth"

>\color{blue}{Typos corrected.}

Mapping
Figure 1: A narrower border on counties would better highlight the fill color.
Figure 2: Rainbow colors are difficult to interpret. A divergent color band would be better.
Figure 3: A single color could display counties with data. This will link the map to the binary variable in the model.

>\color{blue}{Changes in the cartography include narrower borders for the boundaries and the use of different color palettes. I used palettes from the `scico` package. It provides perceptually uniform and colourblind safe palettes.}

>\color{blue}{I wish to thank you again for the care paid to reviewing this submission. I hope that you will find my responses to your concerns to be fair and sufficient.}

## References

### 2021-09-16 R0-Density-Reanalysis/R0-Density-Reanalysis.Rmd
Revisions after decision by Geographical Analysis

---
title: "Reproducibility of research during COVID-19: examining the case of population density and the basic reproductive rate from the perspective of spatial analysis"
author:
  - name: Author
    email: author@institution.edu
    affiliation: University
    footnote: 1
address:
  - code: University
    address: Department, Street, City, State ZIP
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: |
  The emergence of the novel SARS-CoV-2 coronavirus and the global COVID-19 pandemic has led to explosive growth in scientific research. Alas, much of the research in the literature lacks conditions to be reproducible, and recent publications on the association between population density and the basic reproductive number of SARS-CoV-2 are no exception. Relatively few papers share code and data sufficiently, which hinders not only verification but additional experimentation. In this paper, an example of reproducible research shows the potential of spatial analysis for epidemiology research during COVID-19. Transparency and openness means that independent researchers can, with relatively modest efforts, verify findings and use different approaches as appropriate. Given the high stakes of the situation, it is essential that scientific findings, on which good policy depends, are as robust as possible; as the empirical example shows, reproducibility is one of the keys to ensure this.

@@

```{r load-packages, include=FALSE, cache=FALSE}
library(adespatial) # To calculate Moran eigenvector maps
library(censReg) # Package to estimate tobit model
library(kableExtra) # Package for creating tables
library(Matrix) # Used to convert list of spatial weights to sparse matrix; needed for spatialprobit
library(nlme) # Used to estimate 
library(patchwork)
library(r0density) # Data package
library(sampleSelection) # Package to estimate sampling selection models
library(sf) # Package to work with spatial information in simple features format
library(spatialprobit) # Used to estimate spatially autoregressive tobit
library(spdep) # Utilities to create spatial weights and work with spatial data
library(tidycensus) # Package to retrieve census information 
library(tidyverse) # Data carpentry and analysis tools
library(units) # Package to work with units of measurement
library(viridis) # Color palettes
```

@@

Given the ruinous human and economic cost of the pandemic, there has been a natural tension in the scientific community between the need to publish research results quickly and the imperative to maintain consistently high quality standards in scientific reporting; indeed, a call for maintaining the standards in published research has termed the deluge of COVID-19 publications a "carnage of substandard research" [@Bramstedt2020carnage]. Part of the challenge of maintaining quality standards in published research is that, despite an abundance of recommendations and guidelines [e.g., @Broggini2017reproducible; @Ince2012case; @Ioannidis2014increasing; @Brunsdon2020opening], in practice reproducibility has remained a lofty and somewhat aspirational goal [@Konkol2019examination; @Konkol2019computational]. As reported in the literature, only a woefully small proportion of published research was actually reproducible before the pandemic [@Iqbal2016reproducible; @Stodden2018empirical], and the situation does not appear to have changed substantially since [@Sumner2020reproducibility; @Gustot2020quality].

The push for open data and software, along with more strenuous efforts towards open, reproducible research, is simply a continuation of long-standing scientific practices of independent verification. Despite the (at times disproportionate) attention that high profile scandals in science tend to elicit in the media, science as a collective endeavor is remarkable for being a self-correcting enterprise, one with built-in mechanisms and incentives to weed out erroneous ideas. Over the long term, facts tend to prevail in science. At stake is the shorter-term impacts that research may have in other spheres of economic and social life. The case of economists Reinhart and Rogoff comes to mind: by the time the inaccuracies and errors in their research were uncovered [see @Herndon2014high], their claims about debt and economic growth had already been seized by policy-makers on both sides of the Atlantic to justify austerity policies in the aftermath of the Great Recession of 2007-2009^[Nobel Prize in Economics Paul Krugman noted that "Reinhart–Rogoff may have had more immediate influence on public debate than any previous paper in the history of economics" \url{https://www.nybooks.com/articles/2013/06/06/how-case-austerity-has-crumbled/?pagination=false}]. As later research has demonstrated, those policies cast a long shadow, and their sequels continued to be felt for years [@Basu2017ten].

In the context of COVID-19, a topic that has grabbed the imagination of numerous thinkers has been the prospect of life in cities after the pandemic [@Florida2020how]; the implications of the pandemic for urban planning, design, and management are the topic of ongoing research [@Sharifi2020covid]. The fact that the worst of the pandemic was initially felt in dense population centers such as Wuhan, Milan, Madrid, and New York, unleashed a torrent of research into the associations between density and the spread of the pandemic. he answers to some important questions hang on the results of these research efforts. For example, are lower density regions safer from the pandemic? Are de-densification policies warranted, even if just in the short term? And in the longer term, will the risks of life in high density regions presage a flight from cities? What are the implications of the pandemic for future urban planning and practice? Over the past year, numerous papers have sought to throw light on the underlying issue of density and the pandemic; nonetheless the results, as will be detailed next, remain mixed. Further, to complicate matters, precious few of these studies appear to be sufficiently open to support independent verification.

The objective of this paper is to illustrate the importance of reproducibility in research in the context of the flood of COVID-19 papers. To this end, a recent study by Sy et al. [-@Sy2021population] is chosen as an example of reproducible research. The objective is not to malign the analysis of these researchers, but rather to demonstrate the value of openness to allow for independent verification and further analysis. Open data and open code mean that an independent researcher can, with only modest efforts, not only verify the findings reported, but also examine the same data from a perspective which may not have been available to the original researchers due to differences in disciplinary perspectives, methodological traditions, and/or training, among other possible factors. The example, which shows consequential changes in the conclusions reached by different analyses, should serve as a call to researchers to redouble their efforts to increase transparency and reproducibility in research. The present paper, in addition, aims to show how data can be packaged in well-documented, shareable units, and code can be embedded into self-contained documents suitable for review and independent verification. The source for this paper is an [R Markdown](http://rmarkdown.rstudio.com) document which, along with the data package, will be available in a public repository^[For peer-review purposes, the data package and code are currently in an anonymous Drive folder: \url{https://drive.google.com/drive/folders/1cT6tcUc1pJ4aT5ajQ0emO0lyS46P8Ige?usp=sharing}].

@@

The concern with population density and the spread of the virus during the COVID-19 pandemic was fueled, at least in part, by dramatic scenes seen in real-time around the world from large urban centers such as Wuhan, Milan, Madrid, and New York. In theory, there are good reasons to believe that higher density may have a positive association with the transmission of a contagious virus. It has long been known that the potential for inter-personal contact is greater in regions with higher density [see for example the research on urban fields and time-geography, including @Farber2011running; @Moore1970urban; @Moore1970some]. Mathematically, models of exposure and contagion indicate that higher densities can catalyze the transmission of contagious diseases [@Rocklov2020high; @Li2018effect]. The idea is intuitive and likely at the root of messages, by some figures in positions of authority, that regions with sparse population densities faced lower risks from the pandemic^[Governor Kristi Noem of South Dakota, for example, claimed that sparse population density allowed her state to face the pandemic down without the need for strict policy interventions \url{https://www.inforum.com/lifestyle/health/5025620-South-Dakota-is-not-New-York-City-Noem-defends-lack-of-statewide-COVID-19-restrictions}].

As Rocklöv and Sjödin [@Rocklov2020high] note, however, mathematical models of contagion are valid at small-to-medium spatial scales (and presumably, small temporal scales too, such as time spent in restaurants, concert halls, cruises), and the results do not necessarily transfer to larger spatial units and different time scales. There are solid reasons for this: while in a restaurant, one can hardly avoid being in proximity to other customers. On the other hand, a person can choose to (or be forced to as a matter of policy) not go to a restaurant in the first place. Nonetheless, the idea that high density correlates with high transmission is so seemingly sensible that it is often taken for granted even at larger scales [e.g., @Cruz2020exploring; @Micallef2020first]. At larger scales, however, there exists the possibility of behavioral adaptations, which are difficult to capture in the mechanistic framework of differential equations [or can be missing in agent-based models, e.g., @Gomez2021infekta]; these adaptations, in fact, can be a key aspect of disease transmission.

@@

On the flip side, a number of studies report non-significant or negative associations between population density and COVID-19 outcomes. This includes the research of Sun et al. [-@Sun2020impacts] who did not find evidence of significant correlation between population density and confirmed number of cases per day _in conditions of lockdown_ in China. This finding echoes the results of Paez et al. [-@Paez2020spatio], who in their study of provinces in Spain reported non-significant associations between population density and infection rates in the early days of the first wave of COVID-19, and negative significant associations in the later part of the first lockdown. Similarly, [-@Skorka2020macroecology] found zero or negative associations between population density and infection numbers/deaths by country. Fielding-Miller et al. [-@Fielding2020social] contrast their finding about rural counties with a negative relationship between COVID-19 deaths and population density in urban counties in the US. For their part, in their investigation of doubling time, White and Hébert-Dufresne [-@White2020state] identified a negative and significant correlation between population density and doubling time in US states. Likewise, [-@Khavarian2021high] found a small negative (and significant) association between population density and COVID-19 morbidity in districts in Tehran. Finally, two of the most complete studies in the US [by Hamidi et al. -@Hamidi2020longitudinal; and -@Hamidi2020density] used an extensive set of controls to find negative and significant correlations between density and COVID-19 cases and fatalities at the level of counties in the US.

As can be seen, these studies are implemented at different scales in different regions of the world. They also use a range of techniques, from correlation analysis, to multivariate regression, spatial regressions, and machine learning techniques. This is natural and to be expected: individual researchers have only limited time and expertise. This is why reproducibility is important. To pick an example (which will be further elaborated in later sections of this paper), the study of Sy et al. [-@Sy2021population; hereafter SWN] would immediately grab the attention of a researcher with expertise in spatial analysis.

@@

SWN investigated the basic reproductive number of COVID-19 in US counties, and its association with population density, median household income, and prevalence of private mobility. For their multivariate analysis, SWN used mixed linear models. This is a reasonable modelling choice: $R_0$ is an interval-ratio variable that is suitably modeled using linear regression; further, as SWN note there is a likelihood that the process in not independent "among counties within each state, potentially due to variable resource allocation and differing health systems across states"  (p. 3). A mixed linear model accounts for this by introducing random components; in the case of SWN, these are random intercepts at the state level. SWN estimated various models with different combinations of variables, including median household income and prevalence of travel by private transportation. These are sensible controls, given potential variations in behavior: people in more affluent counties may have greater opportunities to work from home, and use of private transportation reduces contact with strangers. Moreover, they also conducted various sensitivity analyses. After these efforts, SWN concluded that there is a positive association between the basic reproductive number and population density at the level of counties in the US.

>>>>

---
title: "Reproducibility of research during COVID-19: examining the case of population density and the basic reproductive rate from the perspective of spatial analysis"
author:
  - name: Author
    email: author@institution.edu
    affiliation: University
    footnote: 1
address:
  - code: University
    address: Department, Street, City, State ZIP
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: |
  The emergence of the novel SARS-CoV-2 coronavirus and the global COVID-19 pandemic in 2019 led to explosive growth in scientific research. Alas, much of the research in the literature lacks conditions to be reproducible, and recent publications on the association between population density and the basic reproductive number of SARS-CoV-2 are no exception. Relatively few papers share code and data sufficiently, which hinders not only verification but additional experimentation. In this paper, an example of reproducible research shows the potential of spatial analysis for epidemiology research during COVID-19. Transparency and openness means that independent researchers can, with only modest efforts, verify findings and use different approaches as appropriate. Given the high stakes of the situation, it is essential that scientific findings, on which good policy depends, are as robust as possible; as the empirical example shows, reproducibility is one of the keys to ensure this.

@@

```{r load-packages, include=FALSE, cache=FALSE}
library(adespatial) # Used to calculate Moran eigenvector maps
library(censReg) # Package to estimate tobit model
library(kableExtra) # Package for creating tables
library(Matrix) # Used to convert list of spatial weights to sparse matrix; needed for spatialprobit
library(nlme) # Used to estimate 
library(patchwork) # Package to organize ggplot2 plots
library(r0density) # Data package
library(sampleSelection) # Package to estimate sampling selection models
library(scico) # Package with color palettes
library(sf) # Package to work with spatial information in simple features format
library(spatialprobit) # Used to estimate spatially autoregressive tobit
library(spdep) # Utilities to create spatial weights and work with spatial data
library(tidycensus) # Package to retrieve census information 
library(tidyverse) # Data carpentry and analysis tools
library(units) # Package to work with units of measurement
#library(viridis) # Color palettes
```

@@

Given the ruinous human and economic cost of the pandemic, there has been a natural tension in the scientific community between the need to publish research results quickly and the imperative to maintain consistently high quality standards in scientific reporting; indeed, a call for maintaining the standards in published research termed the deluge of COVID-19 publications a "carnage of substandard research" [@Bramstedt2020carnage]. Part of the challenge of maintaining quality standards in published research is that, despite an abundance of recommendations and guidelines [e.g., @Broggini2017reproducible; @Ince2012case; @Ioannidis2014increasing; @Brunsdon2020opening], in practice reproducibility has remained a lofty and somewhat aspirational goal [@Konkol2019examination; @Konkol2019computational]. As reported in the literature, only a woefully small proportion of published research was actually reproducible before the pandemic [@Iqbal2016reproducible; @Stodden2018empirical], and the situation does not appear to have changed substantially since [@Sumner2020reproducibility; @Gustot2020quality].

The push for open software and data [e.g., @Bivand2020progress; @Arribas2021open], along with more strenuous efforts towards open, reproducible research, is simply a continuation of long-standing scientific practices of independent verification. Despite the (at times disproportionate) attention that high profile scandals in science tend to elicit in the media, science as a collective endeavor is remarkable for being a self-correcting enterprise, one with built-in mechanisms and incentives to weed out erroneous ideas. Over the long term, facts tend to prevail in science. At stake is the shorter-term impacts that research may have in other spheres of economic and social life. The case of economists Reinhart and Rogoff comes to mind: by the time the inaccuracies and errors in their research were uncovered [see @Herndon2014high], their claims about debt and economic growth had already been seized by policy-makers on both sides of the Atlantic to justify austerity policies in the aftermath of the Great Recession of 2007-2009^[Nobel Prize in Economics Paul Krugman noted that "Reinhart–Rogoff may have had more immediate influence on public debate than any previous paper in the history of economics" \url{https://www.nybooks.com/articles/2013/06/06/how-case-austerity-has-crumbled/?pagination=false}]. As later research has demonstrated, those policies cast a long shadow, and their sequels continued to be felt for years [@Basu2017ten].

In the context of COVID-19, a topic that has grabbed the imagination of numerous thinkers has been the prospect of life in cities after the pandemic [e.g., @Florida2020how]; as a result, the implications of the pandemic for urban planning, design, and management are the topic of ongoing research [e.g., @Sharifi2020covid]. The fact that the worst of the pandemic was initially felt in dense population centers such as Wuhan, Milan, Madrid, and New York, unleashed a torrent of research into the associations between density and the spread of the pandemic. The answers to some important questions hang on the results of these research efforts. For example, are lower density regions safer from the pandemic? Are de-densification policies warranted, even if just in the short term? In the longer term, will the risks of life in high density regions presage a flight from cities? And, what are the implications of the pandemic for future urban planning and practice? Over the past year, numerous papers have sought to throw light on the underlying issue of density and the pandemic; nonetheless the results, as will be detailed next, remain mixed. Further, to complicate matters, precious few of these studies appear to be sufficiently open to support independent verification.

The objective of this paper is to illustrate the importance of reproducibility in research in the context of the flood of COVID-19 papers. For this, I focus on a recent study by Sy et al. [-@Sy2021population] that examined the correlation between the basic reproductive number of COVID-19, $R_0$, and population density.  The basic reproductive number is a summary measure of contact rates, probability of transmission of a pathogen, and duration of infectiousness. In rough terms, it measures how many new infections each infections begets. The paper of Sy et al. [-@Sy2021population] was selected for being, in the literature examined, almost alone in supporting reproducible research. Accordingly, I wish to be clear that my objective in singling their work for discussion is not to malign their efforts, but rather to demonstrate how open and reproducible research efforts can greatly help to accelerate discovery. More concretely, open data and open code mean that an independent researcher can, with only modest efforts, not only verify the findings reported, but also examine the same data from a perspective which may not have been available to the original researchers due to differences in disciplinary perspectives, methodological traditions, and/or training, among other possible factors. The example, which shows consequential changes in the conclusions reached by different analyses, should serve as a call to researchers to redouble their efforts to increase transparency and reproducibility in their research. In this spirit, the present paper also aims to show how data can be packaged in well-documented, shareable units, and code can be embedded into self-contained documents suitable for review and independent verification. The source for this paper is an [R Markdown](http://rmarkdown.rstudio.com) document which, along with the data package, will be available in a public repository^[For peer-review purposes, the data package and code are currently in an anonymous Drive folder: \url{https://drive.google.com/drive/folders/1cT6tcUc1pJ4aT5ajQ0emO0lyS46P8Ige?usp=sharing}].

@@

The concern with population density and the spread of the virus during the COVID-19 pandemic was fueled, at least in part, by dramatic scenes seen in real-time around the world from large urban centers such as Wuhan, Milan, Madrid, and New York. In theory, there are good reasons to believe that higher density could have a positive association with the transmission of a contagious virus. It has long been known that the potential for inter-personal contact is greater in regions with higher density [see for example the research on urban fields and time-geography, including @Farber2011running; @Moore1970urban; @Moore1970some]. Mathematically, models of exposure and contagion indicate that higher densities can catalyze the transmission of contagious diseases [@Rocklov2020high; @Li2018effect]. The idea is intuitive and likely at the root of messages, by some figures in positions of authority, that regions with sparse population densities faced lower risks from the pandemic^[Governor Kristi Noem of South Dakota, for example, claimed that sparse population density allowed her state to face the pandemic down without the need for strict policy interventions \url{https://www.inforum.com/lifestyle/health/5025620-South-Dakota-is-not-New-York-City-Noem-defends-lack-of-statewide-COVID-19-restrictions}].

As Rocklöv and Sjödin [@Rocklov2020high] note, however, mathematical models of contagion are valid at small-to-medium spaces (and presumably, smaller time intervals too, such as time spent in restaurants, concert halls, cruises), and the results do not necessarily transfer to larger spatial units and longer time periods. There are solid reasons for this: while in a restaurant, one can hardly avoid being in proximity to other customers. On the other hand, a person can choose to (or be forced to as a matter of policy) not go to a restaurant in the first place. Nonetheless, the idea that high density correlates with high transmission is so seemingly sensible that it is often taken for granted even at the scale of large spaces [e.g., @Cruz2020exploring; @Micallef2020first]. In such conditions, however, there exists the possibility of behavioral adaptations, which are difficult to capture in the mechanistic framework of differential equations [or can be missing in agent-based models, e.g., @Gomez2021infekta]; these adaptations, in fact, can be a key aspect of disease transmission.

@@

On the flip side, a number of studies report non-significant or negative associations between population density and COVID-19 outcomes. This includes the research of Sun et al. [-@Sun2020impacts] who did not find evidence of significant correlation between population density and confirmed number of cases per day _in conditions of lockdown_ in China. This finding echoes the results of Paez et al. [-@Paez2020spatio], who in their study of provinces in Spain reported non-significant associations between population density and infection rates in the early days of the first wave of COVID-19, and negative significant associations in the later part of the first lockdown. Similarly, @Skorka2020macroecology found zero or negative associations between population density and infection numbers/deaths by country. Fielding-Miller et al. [-@Fielding2020social] contrast their finding about rural counties with a negative relationship between COVID-19 deaths and population density in urban counties in the US. For their part, in their investigation of doubling time, White and Hébert-Dufresne [-@White2020state] identified a negative and significant correlation between population density and doubling time in US states. Likewise, @Khavarian2021high found a small negative (and significant) association between population density and COVID-19 morbidity in districts in Tehran. Finally, two of the most complete studies in the US, by @Hamidi2020longitudinal and @Hamidi2020density, used an extensive set of controls to find negative and significant correlations between density and COVID-19 cases and fatalities at the level of counties in the US.

As can be seen, these studies are implemented at different scales in different regions of the world. They also use a range of techniques, from correlation analysis, to multivariate regression, spatial regressions, and machine learning techniques. This is natural and to be expected: individual researchers have only limited time and expertise. This is why reproducibility is important. To pick an example (which will be further elaborated in later sections of this paper), the study of Sy et al. [-@Sy2021population], hereafter referred to as SWN, would immediately grab the attention of a researcher with expertise in spatial analysis.

@@

SWN investigated the basic reproductive number of COVID-19 in US counties, and its association with population density, median household income, and prevalence of private mobility. For their multivariate analysis, SWN used mixed linear models. This is an appropriate modelling choice: $R_0$ is an interval-ratio variable that is suitably modeled using linear regression; further, as SWN note there is a likelihood that the process in not independent "among counties within each state, potentially due to variable resource allocation and differing health systems across states"  (p. 3). A mixed linear model accounts for this by introducing random components; in the case of SWN, these are random intercepts at the state level. SWN estimated various models with different combinations of variables, including median household income and prevalence of travel by private transportation. These controls help to account for potential variations in behavior: people in more affluent counties may have greater opportunities to work from home, and use of private transportation reduces contact with strangers. Moreover, they also conducted various sensitivity analyses. After these efforts, SWN concluded that there is a positive association between the basic reproductive number and population density at the level of counties in the US.
